\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-07-31 mers-structure.tex
Revision pass

Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.

@@

\section*{Introduction}
%% Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels. % AR - a virus isn't an infection and not all infections in the Arabian Peninsula
Middle East Respiratory Syndrome Coronavirus (MERS-CoV), endemic in camels in the Arabian Peninsula, is the causative agent of zoonotic infections and limited outbreaks in humans. % ? something like this ?
The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{who_2017}.
Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
Whilst contact with camels is often reported, patients may not recall contact with %\lmc{other?}
any livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015}.
Epidemiological modeling approaches have been used to look at traditional data sources of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.

Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches %\lmc{why, specifically?}
due to difficulties in linking cases into clusters in the absence of detailed information.
Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{gire_genomic_2014,quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce.
These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
In the absence of other information often only sequence data can %\lmc{help}
help pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from %\lmc{when you say hospital outbreak, do you mean an actual outbreak taking place inside the hospital?}
outbreaks within hospitals, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines %\lmc{how about a citation here?}
\citep{who_2016_MERS}.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use existing MERS-CoV sequence data (174 MERS-CoV genomes from human infections and 100 MERS-CoV genomes from camel infections) to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of metapopulation structure and migration between discrete sub-populations, referred to here as demes \citep{vaughan_efficient_2014}.

%Unlike discrete trait analyses (DTA, or ``mugration'' models) % AR - I don't like 'DTA' as name for this model - it is not specific enough. The model is a continuous time Markov process
%that treat a potential trait of interest (such as host state) as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly incorporate distinct sampling patterns and population dynamics across demes.
Unlike approaches that model the host species as a discrete trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) %[cite here, Philippe's bat rabies and Sam's avian flu?, others?]
\citep{faria_simultaneously_2013,global_consortium_for_h5n8_and_related_influenza_viruses_role_2016}, population structure models explicitly incorporate the distinct sampling patterns, population dynamics within demes and the migration between demes.
By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
Using Monte Carlo simulations we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.

%\begin{figure}[h]
% \centering
%	\includegraphics[width=0.45\textwidth]{figures/}
%	\caption{\textbf{}
%
%	}
%	\label{}
%\end{figure}

@@

The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
Whilst we %\lmc{estimate?}
estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where owing to poor sampling of MERS-CoV genetic diversity from camels the model cannot dismiss humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly %\lmc{maybe ``non-uniformly'' is redundant here} \gdc{disagree, the point I am trying to make is that the data are mostly human and that human data are further skewed in how they are sampled}
sampled human cases and are robust to choice of prior.
We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
\tbc{I think this sentence (``We therefore...'') needs to be spelled out a bit more. It feels like the logic is what the Monte Carlo sampling does. Either spell out more, or push back to section with figure \ref{ABC}.}

\begin{figure}[h]
 \centering
	\includegraphics[width=0.65\textwidth]{figures/mers_mcc.png}
	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from humans and camels.}
	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered with the structured coalescent.
	Vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
  Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
  Some of the branches leading up to zoonotic transmission of MERS-CoV into humans are too long to have spent much of their time in humans, given our \textit{a priori} belief that MERS-CoV cannot circulate in humans for long periods of time.
  Thus branches encompassing cross-species transmissions are depicted as being in camels right up to the common ancestor or tip of each human outbreak.
	Whilst large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
	}
	\label{mcc}
\end{figure}

\subsection*{Signals of MERS-CoV seasonality}
% Unlike the most simpler %\lmc{simpler?}
% ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny,
% please lets drop 'mugration'. Terrible word. Also I don't see why they are 'simpler'.
% structured population models possess the ability to estimate migration times. % So do Markov Jumps. \gdc{good point, yeah}
We use the posterior distribution of MERS-CoV introduction times of human clades observed in the MCC tree (Figure \ref{mcc}) recovered from structured coalescent to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels resulting in increased force of infection sparking human outbreaks.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_seasonality.png}
	\caption{\textbf{MERS-CoV seasonality.}
A) %Violin plots% %  (the plot type is irrelevant).
Posterior density estimates coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
Posterior means are indicated with circles.
Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
Hatched area spanning October to February indicates the camel calving season.
B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (Figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
The black dotted line indicates the joint probability density for introductions.
%Grey areas spanning October to February indicate the reported camel calving season.\lmc{if these two panels indeed end up being presented together, this is superfluous}
	}
	\label{seasonality}
\end{figure}

%\begin{figure}[h]
% \centering
%	\includegraphics[width=0.45\textwidth]{figures/}
%	\caption{\textbf{}
%
%	}
%	\label{}
%\end{figure}

@@

Structured coalescent approaches clearly show humans to be a terminal host %\tbc{again, I don't know if this is the right term or not}
for MERS-CoV, implying poor transmissibility.
However, we wanted to translate this observation into the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides %\lmc{insights?}
insight into epidemic behavior.
The parameter $R_{0}$, along with the degree of heterogeneity in tranmission quantified by a dispersion parameter $\omega$, can be used to determine the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}).
We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
First, we simulate case counts across multiple outbreaks totalling 2000 cases using Equation \ref{clusterLikelihood} with a fixed dispersion parameter $\omega$ and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing of a fraction of cases.
Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
We record the $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses (left-hand panels in Figure \ref{ABC}).

We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{ABC}).
For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
Whilst the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
Under unbiased sequencing only 0.6\% (2282 out of 363 000) of simulations fit our phylogenetic observations, while 2.7\% and 2.7\%%\lmc{again?}
(nearly 10 000 out of 363 000 for both) of simulations fit for bias levels of 2.0 and 3.0, respectively.
This is apparent from the heights of the histograms in Figure \ref{ABC}, which indicate that a higher number of simulations generate MERS-CoV-like data at bias levels of 2.0 and 3.0 relative to bias of 1.0 (10\% support for bias level 1.0, 45\% support for bias level 2.0, and 45\% support for bias level 3.0).%\lmc{I don't follow this argument...}
Simulations with bias suggest that $R_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).

Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{ABC}).
The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
Our results suggest a large number of unobserved MERS primary cases.%\lmc{does the definition of ``index case'' allow for ``multiple''?} \gdc{not in epi literature I don't think}
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In those sets of analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).

Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling likely to be biased towards large hospital outbreaks.
% \tbc{I didn't understand what you meant by `sequence sampling wholly unrepresentative of viral diversity found in humans'.}
Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.
%\tbc{It would be really nice to be able to reference something here on hospital spread of MERS (or SARS) owing to hygiene practices.}

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_epi.png}
	\caption{\textbf{Results of Monte Carlo simulations.}
Each row corresponds to a different bias value used to concentrate the multivariate hypergeometric distribution used to simulate sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias which make large case clusters to be more likely to be sequenced%\lmc{make sequences from large clusters more likely to be picked}
).
Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the $R_{0}$ value used for the simulation.
The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for MERS-CoV data.
The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same y axis across all levels of bias.
Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same y axis across all levels of bias.
Darker shade of grey indicates bins falling within the 95\% percentiles.
These Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
% \tbc{Not at all essential, but I can imagine that an initial panel A showing the logic of the two step ABC process would do a whole lot to alleviate confusion. I'm picturing a set of tables numbered with case counts in a top row and then a bottom row with sequence counts. This could also work better as a separate figure referenced from the Methods.}
	}
	\label{ABC}
\end{figure}

>>>>

Middle East respiratory syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
While traditional and seroepidemiological studies have been employed extensively throughout the epidemic, viral sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
We employ structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as a transient, and ultimately terminal host.
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that human outbreaks in the Arabian peninsula are driven by seasonally varying zoonotic transfer of viruses from camels.
Without heretofore unseen evolution of host tropism, MERS-CoV is unlikely to become endemic in humans.

@@

\section*{Introduction}
Middle East respiratory syndrome coronavirus (MERS-CoV), endemic in camels in the Arabian Peninsula, is the causative agent of zoonotic infections and limited outbreaks in humans.
The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has caused more than 2000  infections and over 700 deaths, according to the World Health Organization (WHO) \citep{who_2017}.
Its epidemiology remains obscure, largely because infections are observed among the most severely affected individuals, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
While contact with camels is often reported, other patients do not recall contact with any livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
Previous studies on MERS-CoV epidemiology have used serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
Such data have shown high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015}.
Separately, epidemiological modelling approaches have been used to look at incidence reports through time, space and across hosts \citep{cauchemez_unraveling_2016}.

Although such traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much inevitably remains opaque to such approaches due to difficulties in linking cases into transmission clusters in the absence of detailed information.
Genomic epidemiology, however, can fill this critical gap and has repeatedly shown the utility of viral sequence data in outbreak scenarios \citep{liu_h7n9_2013,gire_genomic_2014,grubaugh_multiple_2017}, where data are relatively cheap to produce.
These data can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
Sequence data can help discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from outbreaks within hospitals, where sequence data have been used to determine whether infections were isolated introductions or were part of a larger hospital-associated outbreak \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO guidelines \citep{who_2016_MERS}.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional epidemiological approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use a dataset of 274 MERS-CoV genomes to investigate transmission patterns of the virus between humans and camels.

Here, we use an explicit model of metapopulation structure and migration between discrete subpopulations, referred to here as demes \citep{vaughan_efficient_2014}, derived from the structured coalescent \citep{notohara_structured_coalescent_1990}.
Unlike approaches that model host species as a discrete phylogenetic trait of the virus using continuous-time Markov processes (or simpler, parsimony based, approaches) \citep{faria_simultaneously_2013,lycett_h5n8_2016}, population structure models explicitly incorporate distinct sampling patterns, population dynamics within demes and migration between demes.
By estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
Using Monte Carlo simulations we show that $R_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that correspondingly the virus has been introduced into humans hundreds of times.

@@

The structured coalescent approach we employ (see Methods) identifies camels as a reservoir host where most of MERS-CoV evolution takes place (Figure \ref{mcc}), while human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (Figure~\ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we estimate a median of 56 separate camel-to-human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
While we estimate a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (Figure~\ref{mcc}).
This inference derives from the tree structure wherein human viruses appear as clusters of highly related sequences nested within the diversity seen in camel viruses, which themselves show signicantly higher diversity and less clustering.
This manifests as different rates of coalescence with camel viruses showing a scaled effective population size $N_e \tau$ of \tbc{XXX (95\% HPD: XXX--XXX)} and human viruses showing a scaled effective population of \tbc{XXX (95\% HPD: XXX--XXX)}.

We believe that the small number of inferred human-to-camel migrations are induced by the migration rate prior, while some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
This is most apparent in lineages in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where owing to poor sampling of MERS-CoV genetic diversity from camels the model cannot completely dismiss humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (Figure~\ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
This suggests that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{figures/mers_mcc.png}
	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from 174 human viruses and 100 camel viruses.}
	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered with the structured coalescent.
	The vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
  Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
  Some of the branches leading up to zoonotic transmission of MERS-CoV into humans are too long to have spent much of their time in humans, given our \textit{a priori} belief that MERS-CoV cannot circulate in humans for long periods of time.
  Thus branches encompassing cross-species transmissions are depicted as being in camels right up to the common ancestor or tip of each human outbreak.
  \tbc{The previous two sentences (``Some of the...'') seem unnecessary and confusingly written. I would just drop them.}
	While large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
	}
	\label{mcc}
\end{figure}

\subsection*{MERS-CoV shows seasonal introductions}
We use the posterior distribution of MERS-CoV introduction events from camels to humans (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached \citep{martinez_seasonality_2014}, leading to a sharp rise in prevalence of the virus in camels and resulting in increased force of infection into the human population.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

\tbc{I think this is the best place to discuss the lack of correlation between date and size of human cluster. This can be referenced later on in Discussion when talking about pandemic potential. Also, seems relevant to look at seasonality in size of cluster here, yes? This would get at whether there are seasonal factors on the human side.}

\begin{figure}[h]
\centering
	\includegraphics[width=0.75\textwidth]{figures/mers_seasonality.png}
	\caption{\textbf{Seasonality of MERS-CoV introduction events.}
A) Posterior density estimates partitioned by month showing the 95\% highest posterior density interval for relative odds ratios of MERS-CoV introductions into humans.
Posterior means are indicated with circles.
Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
Hatched area spanning October to February indicates the camel calving season.
B) Sequence cluster sizes and inferred dates of introduction events.
Each introduction event is shown as a vertical line positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendant sequences recovered from human cases.
95\% highest posterior density intervals for introductions of MERS-CoV into humans are indicated with coloured lines, coloured by median estimated introduction time.
The black dotted line indicates the joint probability density for introductions.
We find little correlation between date and size of introduction (Spearman $\rho = 0.06$, $p=0.68$).
	}
	\label{seasonality}
\end{figure}

@@

Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
However, we wanted to translate this observation into an estimate of the basic reproductive number, $R_{0}$, which is more familiar to epidemiologists and provides insight into epidemic behavior.
The parameter $R_{0}$ determines expected number of secondary cases in a single infections as well as the distribution of total cases that result from a single introduction event into the human population (Equation~\ref{clusterLikelihood}, Methods).
We estimate $R_{0}$ along with other relevant parameters via Monte Carlo simulation in two steps.
First, we simulate case counts across multiple outbreaks totaling 2000 cases using Equation \ref{clusterLikelihood} and then we subsample from each case cluster to simulate sequencing of a fraction of cases.
Sequencing simulations take place at different levels of bias, wherein bias enriches sequencing of larger case clusters.
This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from primary cases will be sampled exceedingly rarely.
We record the mean, median and standard deviation of sequence cluster sizes in each simulation (left-hand panels in Figure \ref{mers_epi}) and extract the subset of Monte Carlo simulations in which these summary statistics fall within the 95\% highest posterior density observed in the empirical MERS-CoV data from structured coalescent analyses.
We record $R_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions), for these empirically matched simulations.
A schematic of this Monte Carlo produre is shown in Figure \ref{mc_method}.
Generally, higher $R_0$ results in fewer larger transmission clusters, while lower $R_0$ results in many smaller transmission clusters (Figure \ref{mers_epi}).

We find that observed phylogenetic patterns of sequence clustering strongly support $R_{0}$ below 1.0 (middle panels in Figure \ref{mers_epi}).
For increasing levels of bias mean $R_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
While the 95\% percentiles for $R_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.}\ uniform sequencing efforts, in which every case is equally likely to be sequenced), we also note that increasing levels of bias are considerably more to likely to generate MERS-CoV-like sequence clusters (Figure \ref{mers_epi}).
Under unbiased sequencing only 0.6\% of simulations fit our phylogenetic observations, while 2.7\% and 2.7\% of simulations fit for bias levels of 2.0 and 3.0, respectively.
Correspondingly, we estimate 10\% support for a model with bias level 1.0, 45\% support for a model with bias level 2.0, and 45\% support for a model with bias level 3.0.
Model averaging would suggest plausible $R_0$ values between \tbc{XXX} and \tbc{XXX}.

Lower values for $R_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans (right-hand panels in Figure \ref{mers_epi}).
The median number of cross-species introductions observed in simulations matching empirical data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously match considerably better to MERS phylogenetic data.
Model averaging would suggest plausible numbers of introductions between \tbc{XXX} and \tbc{XXX}.
Our results suggest a large number of unobserved MERS primary cases.
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In these analyses $R_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (Figure \ref{extra_cases}).

Overall, our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated $R_{0}<1.0$ and sequence sampling likely to be biased towards large hospital outbreaks.
Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices \citep{assiri_hospital_2013,chen_comparative_2017} appear sufficient to explain observed epidemiological patterns of MERS-CoV.

\begin{figure}[h]
\centering
	\includegraphics[width=0.75\textwidth]{figures/mers_epi.png}
	\caption{\textbf{Monte Carlo simulations of human transmission clusters.}
Each row corresponds to a different bias value used to concentrate the subsampling of sequences from cases, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias which make large case clusters to be more likely to be sequenced).
Leftmost scatter plots show results of individual Monte Carlo simulations, coloured by the $R_{0}$ value used for the simulation.
The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for empirical MERS-CoV data.
The distribution of $R_{0}$ values found within the dotted rectangle is shown in the middle, on the same $y$-axis across all levels of bias.
Bins falling inside the 95\% percentiles are coloured by $R_{0}$, as in the leftmost scatter plot.
The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same $y$-axis across all levels of bias.
Darker shade of grey indicates bins falling within the 95\% percentiles.
These Monte Carlo simulations indicate $R_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
	}
	\label{mers_epi}
\end{figure}
